---
output:
  pdf_document: default
  html_document: default
---
# Data Science: The Tools of the Trade{#intro}

Using the right set of tools in a workflow that's structured correctly will make you a much more effective data analyst. In our eperience, the effectiveness of the tools of data science are enhanced by orders of magnitude when using an appropriate setup. Like chefs who structure their "mise en scene" or woodworkers who properly organize their shop, the right structure for doing your work can make all the difference. 

In this chapter we'll be working on getting you set up with the tools you will need to undertake complex data analysis projects. Once the tools are in place, we'll take a look at the basic building block of coding in R. 


## Overview

We need three basic sets of tools for this class. We will need `R` to analyze data. We will need `RStudio` to help us interface with R and to produce documentation of our results. Last, we will need `git` and GitHub to communicate our results to the wider world.  

## Installing R

R is an extensible statistical programming environment that can handle all of the main tasks of data science: getting data, analyzing data and communicating data analysis. As a bonus, R is also free and open source, and has a large and vibrant community of collaborators who are constantly engaging one another to build better tools. 

You can download R here: [https://cran.r-project.org/](https://cran.r-project.org/). 

Once R is downloaded, you should run it from your computer. You should get a program window that looks like this:

As you can see, the interface for R is pretty basic. You can type in commands at the prompt and it will give you some output. For example, you could type `2+2` at the prompt, and (hopefullyu) the answer 4 will be displayed. R uses what's called a command line interface, which is quite different than the "point and click" approach that is typically used when we interact with computers. To work fully with the capabilities of R, we'll need another set of tools. 

## Installing Rstudio

When we work with R, we communicate via the command line. To help automate this process, we can write scripts, which contain all of the commands to be executed. These scripts generate various kinds of output, like numbers on the screen, graphics or reports in common formats (pdf, word). Most programming languages have several __I__ ntegrated __D__ evelopment __E__ nvironments (IDEs) that encompass all of these elements (scripts, command line interface, output). The primary IDE for R is Rstudio. 

If you haven't already, you need to download Rstudio here:
[https://www.rstudio.com/products/rstudio/download2/](https://www.rstudio.com/products/rstudio/download2/). You need the free Rstudio desktop version. Once you've downloaded Rstudio, you should run it, and see a window that looks something like this:

There's a lot there that we need to go through, but for now just having gotten this up and running will be enough. 

## Installing git

`git` is a version control program. A standard problem in programming is how to track the changes that have been made to a project. These challenges are well-known to anyone who has had to work on a complex document, particularly with collaborators. Because these problems are particularly severe in programming, programmers developed a set of tools called version control. Version control will keep track of changes you make to a file, and record why you made the change. There are many more things that version control can do, but the basics of tracking what changed and why will be our focus to start. 

Download git here and accept all of the defaults on installation: https://git-scm.com/downloads

There's not really a program to run to see git on your computer, but we'll check that it's working in the next sections. 

## Getting set up on GitHub

GitHub is online hosting service that is widely used by programmers. It allows you to easily share your work with the world. If you want to be a member of the data science community, it pays to be engaged with GitHub. GitHub is integrated with git, so version control is easy to do. You should go to: https://github.com and sign up for free to get a username. 

## Initializing repos

Go to your github page and create a new repository. It will ask you if you'd like to create a README file, and you should say yes. Name the repository something short and easy to remember, and initialize it. 

## Installing Github Desktop

Github desktop is a __G__ raphical __U__ ser __I__ nterface for git. We won't use it much, but we will use it to copy your new repository 

Download github desktop: https://desktop.github.com/. Choose the beta version, which appears to be working nicely. Accept all of the defaults on installation. 

Unzip and open up the github app. Then using the app, sign into github using your username and ID. 

## Integrating Rstudio with GitHub

To integrate Rstudio with GitHub, you need to enable git as your version control within Rstudio. Go to Preferences-->Git/SVN and make sure the "enable version control" box is checked. Make sure that you can see a path to git. 

Further instructions (if needed) are [here](https://support.rstudio.com/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN). 

Then you'll need to generate an SSH key. This will let GitHub know it's safe to talk to your computer. In RStudio, go to Preferences--> Git/SVN, then click on "Create RSA Key". Once that's done, you can click on the blue text that says "View public key." Copy all of the text in the public key.

Now go to https://github.com and in the upper right hand corner by your avatar, click on Settings. Go to SSH and GPG keys. Click "New SSH Key" and paste in the text you copied from Rstudio. Name the key, and you should be all set.  

Now, go to your repo  Click the green button that says "clone or download," then copy the link provided by clicking on the clipboard next to the link. 

Now go to Github desktop, and click the "clone a repository." Paste in the link you just copied, choose a location on your computer, and proceed. Choose a good spot on your computer for the project files-- wherever you keep important work. This project will be the only place you need to work as you're learning data science.  It should show you a link to your repository. You now have your local repository (on your computer) linked to your account on GitHub. 

## Linking Github and Rstudo
From RStudio, you'll need to click New Project--Existing Directory-- then choose the directory where you just downloaded the GitHub repository. 

Name the project something short and easy to remember (the same as your GitHub repo above is a good option).  Once you click "create project," you should see a "git" tab in your environment. Open up the file named `01-intro.Rmd` and take a look.  

## Yes We Code! Running R Code

When you open up the `01-intro.Rmd` file you should see something that looks like this: 

This is 

The following code chunk will be our first use of R in this class. We're going to grab some data that's part of the [college scorecard](https://collegescorecard.ed.gov/data/documentation/) and do a bit of analysis on it. 


## .Rmd files

Open the `01-Intro.Rmd` file. In Rstudio, go to File-->Open, then find the `01-Intro.Rmd` file in the directory. 

.Rmd files will be the only file format we work in this class. .Rmd files contain two basic elements:

1. Script that can be interpreted by R. 
2. Text that can be read by humans. 

From a .Rmd file you can generate html documents, pdf documents, word documents, slides . . . lots of stuff. All class notes will be in .Rmd. All assignments will be turned in as .Rmd files, and your final project? You guessed it, .Rmd. 

In the `01-Intro.Rmd` file you'll notice that there are three open single quotes in a row, like so: ` ``` ` This indicates the start of a "code chunk" in our file. The first code chunk that we load will include a set of programs that we will need all semester long.  

When we say that R is extensible, we mean that people in the community can write programs that everyone else can use. These are called "packages." In these first few lines of code, I load a set of packages using the library command in R. The set of packages, called `tidyverse` were written by Hadley Wickham and play a key role in his book.  To install this set of packages, simply type in `install.packages("tidyverse")` at the R command prompt. 

To run the code below in R, you can:

- Press the "play" button next to the code chunk
- In OS X, place the cursor inside the code chunk and hit Command+Enter. In Windows, hit Control+Enter
```{r scorecard} 
## Clear environment
rm(list=ls())
## Get necessary libraries-- won't work the first time, because you need to install them!
library(tidyverse)
```


## Loading Data
Now we're ready to load in data. The data frame will be our basic way of interacting with everything data-related in this class. The `sc` data frame contains information from the college scorecard on 127 different colleges and univeristies.  

However, we first need to make sure that R is looking in the right place. When you opened up your project, Rstudio automagically took you to the directory for that project, so you should be ok, but we should make sure to point R to the right place. This is called setting the working directory, and can be done either by using the command `setwd` or in RStudio by going to "Session-->Set Working Directory-->Choose Directory." Choose the directory where the file currently resides on your computer. Make sure to always set the working directory at the beginning of each session---not doing so causes a lot of headaches for new users. 

```{r load_data}
## Load in the data
load("college.Rdata")
```
Here are the variables in the `college.Rdata` dataset:

*Variable Name* :*Definition*
unitid: Unit ID

instnm: Institution Name

stabbr: State Abbreviation

year: Year

control: control of institution, 1=public, 2= private non-profit, 3=private for-profit

preddeg: predominant degree, 1= certificate, 2= associates, 3= bachelor's, 4=graduate

adm_rate: Proportion of Applicants Admitted

sat_avg: Midpoint of entrance exam scores, on SAT scale, math and verbal only

costt_4a: Average cost of attendance (tuition and room and board less all grant aid)

debt_mdn: Median debt of graduates

md_earn_ne_pg: Earnings of graduates who are not enrolled in higher education, six years after graduation

## Looking at datasets

We can look at the first few rows and columns of `sc` by typing in the data name. 
 
```{r look_at_data}
sc
```

The name "sc" is arbitrary (it's called that because it's ScoreCard data). It simply refers to an object in the curent environment which happens to be a data frame.

We can look at the whole dataset using View.
```{r}
#View(sc)
```


You'll notice that this data is arranged in a rectangular format, with each row showing a different college, and each column representing a different characteristic of that college. Datasets are always structured this way--- cases (or units) will form the rows, and the characteristics of those cases-- or variables--- will form the columns. Unlike working with spreadsheets, this structure is always assumed for datasets. 


## Filter, Select, Arrange

In exploring data, many times we want to look at smaller parts of the dataset. There are three commands we'll use today that help with this. 

-`filter` selects only those cases or rows  that meet some logical criteria. 

-`select` selects only those variables or coloumns that meet some criteria

-`arrange` arranges the rows of a dataset in the way we want. 

For more on these, please see this [vignette](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html).

Let's grab just the data for Vanderbilt, then look only at the average test scores and admit rate. We can use filter to look at all of the variables for Vanderbilt:

```{r filter_select_arrange}
## Where are we?
sc%>%
  filter(instnm=="Vanderbilt University")
```

Many times, though we don't want to see everything, we just want to choose a few variables. `select` allows us to select only the variables we want. In this case, the institution name, its admit rate, and the average SAT scores of entering students. 
```{r}
sc%>%
  filter(instnm=="Vanderbilt University")%>%
  select(instnm,adm_rate,sat_avg )
```

`filter` takes logical tests as its argument. The code `insntnm=="Vanderbilt University"` is a logical statement that will be true of just one case in the dataset-- when institution name is Vanderbilt University. The `==` is a logical test, asking if this is equal to that. Other common logical and relational operators for R include

-`>`, `<`: greater than, less than
-`>=`, `<=`: greater than or equal to, less than or equal to
- `!` :not, as in `!=` not equal to 
- `&` AND
- `|` OR

Next, we can use `filter` to look at colleges with low admissions rates, say less than 10% ( or .1 in the proportion scale used in the dataset). 

```{r}
sc%>%
  filter(adm_rate<.1)%>%
  select(instnm,adm_rate,sat_avg)%>%
  arrange(sat_avg,adm_rate)
```


Now let's look at colleges with low admit rates, and then order the results by sat scores. Putting the negative sign in front of `sat_avg` gives the results in descending order (`-sat_avg`).

```{r}
sc%>%
  filter(adm_rate<.1)%>%
  select(instnm,adm_rate,sat_avg)%>%
  arrange(-sat_avg)
```

And one last operation: all colleges that admit between 20 and 30 percent of students, looking at their SAT scores, earnings of attendees six years letter, and what state they are in, then arranging by state, and then SAT score. 
```{r}
sc%>%
  filter(adm_rate>.2&adm_rate<.3)%>%
  select(instnm,sat_avg,md_earn_wne_p6,stabbr)%>%
  arrange(stabbr,-sat_avg)%>%
  print(n=20)
```

*Quick Exercise*
Choose a different college and two different things about that college. 

## Summarizing Data*

To summarize data, we use the summarize command. Inside that command, we tell R two things: what to call the new object (a data frame, really) that we're creating, and what numerical summary we would like. The code below summarizes median debt for the colleges in the dataset by calculating the average of median debt for all institutions.

```{r summarize}
sc%>%
  summarize(mean_debt=mean(debt_mdn,na.rm=TRUE))
```

*Quick Exercise*
Summarize the average entering SAT scores in this dataset. 

## Combining Commands
We can also combine commands, so that summaries are done on only a part of the dataset. Below, we summarize median debt for selective schools, and not very selective schools.  

```{r combining commands}
sc%>%
  filter(adm_rate<.1)%>%
  summarize(mean_debt=mean(debt_mdn,na.rm=TRUE))
```

What about for not very selective schools?
```{r}
sc%>%
  filter(adm_rate>.3)%>%
  summarize(mean_debt=mean(debt_mdn,na.rm=TRUE))
```

*Quick Exercise* Calculate average earnings for schools where SAT>1200

*Grouping Data*
Another powerful tool is being able to calculate characteristics for various groups. For example, what are the average earnings for the three different types of colleges (public, private non-profit, private for-profit) in the dataset?

```{r grouping}
sc%>%
  group_by(control)%>%
  summarize(mean_earnings=mean(md_earn_wne_p6))
```
 
What about debt by type of college?  
```{r}
sc%>%
  group_by(control)%>%
  summarize(mean_debt=mean(debt_mdn))
```

*Quick exercise*  Calculate average admission rate by type of college. 

*Plotting Data*
The last basic tool for looking at a dataset is plotting the data. The code below creates a scatterplot of admission rates by average SAT scores

```{r plotting}
gg<-ggplot(data=sc,aes(x=sat_avg,y=adm_rate))
gg<-gg+geom_point()
gg<-gg+geom_smooth()
gg
```

*Quick exercise* Replicate the above plots, but put cost of attendance `costt4_a` on the y axis. 

## Github: save, stage, commit, push

When working with files in your directory, there are several basic steps you can take when using Git as your version control. 

- You should pull at the beginning of every session. You can do this by clicking pull--the down arrrow-- in the upper right hand corner of the git tab in RStudio.

- Saving the file means it is only available to you, on your computer. You should save files as you are working on them. 

- Staging the file means that you would like git (and github) to keep track of the changes you're making to the file. Stage the file by clicking the "staged" box next to it in the Git tab in Rstudio.

- Committing means that you would like to keep the version of the file you generated. Think of this like saving a new version of the file. It doesn't mean that it's done, but it does mean it's in a state that you would like to have a record of as you move forward. To commit a file in Rstudio, click the "commit" button. You will be prompted to add a commit message. There's actually quite a lot of thought about what goes into a commit message. The best idea for now is simply to state why you did what you did, and avoid profanity or any demeaning language. 

- Pushing means that you will send the file and the record of all of the changes made to the file to GitHub. Think of this like emailing it to group members. To push, click the "push" button in the Git tab in Rstudio.  You should do this everytime you finish a working session, at an absolute minimum. 

So every session should go like this:

pull-->do work-->save-->stage(if new file)-->commit

If you'd like you can also use terminal commands to accomplish the same things. Many people (including us) find terminal commands easier to work with. 

- Pull: `git pull`
- Stage: `git add <filenames>`
- Committ: `git commit -m "<your message here>"`
- Push: git push

By far the most useful guide to working with R and git/GitHub is [https://happygitwithr.com/](Jenny Bryan's guide).

## Exercises:

Create a file called `01-assignment_<lastname>.Rmd` in your GitHub repo for assignments. It should contain the following elements:

1. A sentence that says "Hello, World"

2. R output that summarizes one of the variables in the `colllege.Rdata` dataset
3. R output that shows a scatterplot for two of the variables in the `college.Rdata` dataset. 

*Stretch Items*

4. Calculate the average earnings for individuals at the most selective colleges, then compare that with individuals at the least selective colleges in the dataset. 

5. Find a way to determine whether colleges with very high SAT scores tend to be larger or smaller than colleges with low SAT scores. 

6. Plot the relationship between cost and debt. What do you see? Does this surprise you?

7. Now, provide separate plots for cost and debt by control of the institution. 

 Submit it under assignments, using the format `01-assignment_<lastname>.Rmd`. All assignments should be turned in using this format. 

Save, commit and push the assignment file to the assignments directory. 

