# Welcome to Data Science! {#intro}

Today, we'll be working on getting you set up with the tools you will need for this class. Once you are set up, we'll do what we're here to do: analyze data!

Here's what we need to get done today:

1. Introductions  
2. Installing R
3. Installing Rstudio
4. Installing git
5. Getting set up on GitHub
6. Initializing everyone's github repos
7. hello_world.Rmd

## Introductions

## Installing R

R is going to be the only programming language we will use. R is an extensible statistical programming environment that can handle all of the main tasks that we'll need to cover this semester: getting data, analyzing data and communicating data analysis. 

If you haven't already, you need to download R here: [https://cran.r-project.org/](https://cran.r-project.org/). 

## Installing Rstudio

When we work with R, we communicate via the command line. To help automate this process, we can write scripts, which contain all of the commands to be executed. These scripts generate various kinds of output, like numbers on the screen, graphics or reports in common formats (pdf, word). Most programming languages have several _I_ntegrated _D_evelopment _E_nvironments (IDEs) that encompass all of these elements (scripts, command line interface, output). The primary IDE for R is Rstudio. 

If you haven't already, you need to download Rstudio here:
[https://www.rstudio.com/products/rstudio/download2/](https://www.rstudio.com/products/rstudio/download2/). You need the free Rstudio desktop version. 

## Installing git

git is a version control program. A standard problem in programming is how to track the changes that have been made to a project. These challenges are well-known to anyone who has had to work on a complex document, particularly with collaborators. Because these problems are particularly severe in programming, they developed a set of tools called version control. These will keep track of every change you make to a file, and record why you made the change. 

If you haven't already, you need to download git [here](http://git-scm.com/downloads). 

Once you download and install git, you need to introduce yourself to git. Open up a terminal (type in "terminal" in spotlight search in Mac OS X, hit Windows-R and then type in "cmd" in windows) and enter the following two commands:

```
git config --global user.name "YOUR FULL NAME"
git config --global user.email "YOUR EMAIL ADDRESS"
```
Make sure the email address is the same one you use to sign up for GitHub below. 

If you're having trouble installing git, there are a couple of other options.

1. If you're on a Mac, you can install [XCode](https://itunes.apple.com/us/app/xcode/id497799835?mt=12). Git comes bundled with this, and should install cleanly. *N.B.: this is a BIG file and install*

2. Install Github Desktop, available here:[https://desktop.github.com/](https://desktop.github.com/). 

With either, make sure you register your username and email. 

Again on a Mac, if you're having issues with the git path, you may need to do a couple of things. First, go to the terminal and type in 
```
which git
```
This will give you the path to your git executable.

Then, in Rstudio, click "browse" to get to the executable. To be able to find all hidden files, click COMMAND+SHIFT+Period. If it's located somewhere like `/user/local/git/bin/git` you'll need to do something like this. 

You may also want to change the Path, by typing in the following at the terminal:
```
export PATH="$PATH:<path_to_git>""
```

*N.B. in this class, and generally in programming, when you see* `<text>` *that means that you need to substitute something in.*


## Getting set up on GitHub

GitHub is online hosting service that is widely used by programmers. It allows you to easily share your work with the world. It is integrated with git, so version control is easy to do. 

## Initializing repos

Everyone in the class will need a repository (repo from now on) on our GitHub organization. All I need from you is your username on GitHub, then I can add your repo to our organization.

To integrate Rstudio with GitHub, you need to enable git as your version control within Rstudio. 

Instructions are [here](https://support.rstudio.com/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN). 

## Cloning down your repo

First in RStudio, you'll need to generate an SSH key. This will let GitHub know it's safe to talk to your computer. In RStudio, go to Preferences--> Git/SVN, then click on "Create RSA Key". Once that's done, you can click on the blue text that says "View public key." Copy all of the text in the public key.

Now go to GitHub and in the upper right hand corner by your avatar, click on Settings. Go to SSH and GPG keys. Click "New SSH Key" and paste in the text you copied from Rstudio. Name the key, and you should be all set.  

From RStudio, you'll need to click New Project-->Version Control-->GitHub-->Then find your repo, which will be here `https://github.com/hoddatascifall16/student_<yourlastname>`

Name the project "central". Choose a good spot on your computer for the project files-- wherever you usually keep class directories. This project will be the only place you need to work for this class the entire semester. 

Once you click "create project" all of the files should download and you should have your own version of all of the class files downloaded on your computer. 

*N.B. in this class, and generally in programming, when you see* `<text>` *that means that you need to substitute something in.*

## Yes We Code! Running R Code

The following code chunk will be our first use of R in this class. We're going to grab some data that's part of the [college scorecard](https://collegescorecard.ed.gov/data/documentation/) and do a bit of analysis on it. 

When we say that R is extensible, we mean that people in the community can write programs that everyone else can use. These are called "packages." In these first few lines of code, I load two packages using the library command in R. The two packages, `dplyr` and `ggplot2` were written by Hadley Wickham and play a key role in his book. 
```{r scorecard} 
## Clear environment
rm(list=ls())
## Get necessary libraries-- won't work the first time, because you need to install them!
library(tidyverse)
```

Now we're ready to load in data. The data frame will be our basic way of interacting with everything in this class. The `sc` data frame contains information from the college scorecard on 127 different colleges and univeristies. 

```{r load_data}
## Load in the data
load("college.Rdata")
```


Here are the variables in the `college.Rdata` dataset:

*Variable Name* :*Definition*
unitid: Unit ID

instnm: Institution Name

stabbr: State Abbreviation

year: Year

control: control of institution, 1=public, 2= private non-profit, 3=private for-profit

preddeg: predominant degree, 1= certificate, 2= associates, 3= bachelor's, 4=graduate

adm_rate: Proportion of Applicants Admitted

sat_avg: Midpoint of entrance exam scores, on SAT scale, math and verbal only

costt_4a: Average cost of attendance (tuition and room and board less all grant aid)

debt_mdn: Median debt of graduates

md_earn_ne_pg: Earnings of graduates who are not enrolled in higher education, six years after graduation

*Looking at datasets*

We can look at the first few rows and columns of `sc` by typing in the data name. 

We can look at the whole dataset using View. 
```{r look_at_data}
## What does this data look like? Look at the first few rows, first few variables
sc

#View(sc)
```

*Filter, Select, Arrange*

In exploring data, many times we want to look at smaller parts of the dataset. There are three commands we'll use today that help with this. 

-`filter` selects only those cases or rows  that meet some logical criteria. 

-`select` selects only those variables or coloumns that meet some criteria

-`arrange` arranges the rows of a dataset in the way we want. 

For more on these, please see this [vignette](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html).

Let's grab just the data for Vanderbilt, then look only at the average test scores and admit rate. 

```{r filter_select_arrange}
## Where are we?
sc%>%filter(instnm=="Vanderbilt University")

sc%>%filter(instnm=="Vanderbilt University")%>%select(instnm,adm_rate,sat_avg )

## Just colleges with low admit rates: show admit rate and sat scores, arrange in a pleasing way
sc%>%filter(adm_rate<.1)%>%select(instnm,adm_rate,sat_avg)%>%arrange(sat_avg,adm_rate)

## Just colleges with low admit rates: order by sat scores (- sat_avg gives descending)
sc%>%filter(adm_rate<.1)%>%select(instnm,adm_rate,sat_avg)%>%arrange(-sat_avg)
  
## New cut
sc%>%filter(adm_rate>.3)%>%select(instnm,sat_avg,md_earn_wne_p6,stabbr)%>%arrange(stabbr,-sat_avg)%>%print(n=100)
```

*Quick Exercise*
Choose a different college and two different things about that college. 

*Summarizing Data*

```{r summarize}
## What's the average median debt?
sc%>%summarize(mean_debt=mean(debt_mdn,na.rm=TRUE))
```

*Quick Exercise*
Summarize the average entering SAT scores in this dataset. 
```{r}
sc%>%summarize(mean_sat=mean(sat_avg,na.rm=TRUE))
```

*Combining Commands*
We can also combine commands, so that summaries are done on only a part of the dataset. Below, I summarize mediand debt for selective schools, and not very selective schools.  

```{r combining commands}
## What's the average median debt for very selective schools?
sc%>%filter(adm_rate<.1)%>%summarize(mean_debt=mean(debt_mdn,na.rm=TRUE))

## And for not very selective schools?
sc%>%filter(adm_rate>.3)%>%summarize(mean_debt=mean(debt_mdn,na.rm=TRUE))

```

*Quick Exercise* Calcualte average earnings for schools where SAT>1200

```{r quick_sat}
sc%>%filter(sat_avg>1200)%>%summarize(mean_earn=mean(md_earn_wne_p6,na.rm=TRUE))
```

*Grouping Data*
Another powerful tool is being able to calculate characteristics for various groups. For example, what are the average earnings for the three different types of colleges (public, private non-profit, private for-profit) in the dataset?

```{r grouping}
# 
# sc%>%group_by(control)
# %>%
#   summarize(mean_earnings=mean(md_earn_wne_p6))
# 
# sc%>%group_by(control)%>%
#   summarize(mean_debt=mean(debt_mdn))

```

*Plotting Data*
The last basic tool for looking at a dataset is plotting the data. 

```{r plotting}
## Plotting: bivariate
gg<-ggplot(data=sc,aes(x=sat_avg,y=adm_rate))
gg<-gg+geom_point()
gg<-gg+geom_smooth()
gg

## Univariate descriptives

gg<-ggplot(data=sc,aes(adm_rate))
gg<-gg+geom_density()
gg
```

*Quick exercise* Replicate the above plots, but put cost of attendance on the y axis. 

## Your first commit: hello_world.Rmd

.Rmd files will be the only file format we work in this class. .Rmd files contain three basic elements:

1. Script that can be interpreted by R. 
2. Output generated by R, including tables and figures.  
3. Text that can be read by humans. 

From a .Rmd file you can generate html documents, pdf documents, word documents, slides . . . lots of stuff. All class notes will be in .Rmd. All assignments will be turned in as .Rmd files, and your final project? You guessed it, .Rmd. 

For today, I want you to create a file called `hello_world_<lastname>.Rmd` in your github repo for assignment 1. It should contain the following elements:

1. A sentence that says "Hello, World"
2. R output that summarizes one of the variables in the `colllege.Rdata` dataset
3. R output that shows a scatterplot for two of the variables in the `college.Rdata` dataset. 
Lucky for you this is is also your first assignment! Submit it under assignment 1, using the format `assignment1_<lastname>.Rmd`. All assignments should be turned in using this format. 

*Stretch Items*

If you have extra time, you can do the following:

1. Calculate the average earnings for individuals at the most selective colleges, then compare that with individuals at the least selective colleges in the dataset. 
2. Find a way to determine whether colleges with very high SAT scores tend to be larger or smaller than colleges with low SAT scores. 
3. Plot the relationship between cost and debt. What do you see? Does this surprise you? 
4. Now, provide separate plots for cost and debt by control of the institution. 








