---
title: "Small Multiples"
author: "Will Doyle"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=FALSE,fig.width=6.5,fig.height=9)
```

The concept of small multiples was popularized by Edward Tufte in his book, [Envisioning Information](https://www.edwardtufte.com/tufte/books_ei). As Tufte writes: 

"Information slices are positioned within the eyespan, so that viewers make comparisons at a glance---uninterrupted visual reasoning. Constancy of design puts the mephasis on changes in data, not changes in data frames."

If there were one technique I would want people to use more  often in presenting data visually it would be small multiples. A huge amount of information can be conveyed via small multiples, allowing for clear communication of quantitative reasoning.

## Data 

Today we'll be working with data from the World Bank's [World Development Indicators]( http://data.worldbank.org/data-catalog/world-development-indicators) database. The world bank has helpfully created an API, which has an R library built around it, [WDI](https://cran.r-project.org/web/packages/WDI/index.html). 

```{r}

#libraries: WDI interacts with API
library(WDI)
library(tidyverse)
library(forcats)
library(ggplot2)
library(ggthemes)
library(plotly)
```


The WDI library allows us to use the [WDI API](https://datahelpdesk.worldbank.org/knowledgebase/topics/125589) to access data from their database. We're going to start be getting basic country information, including country code, name and region from the built in `WDI_data`.

```{r}

data(WDI_data)

#Pull country info
wd.desc<-data.frame(WDI_data$country)

# Do a quick search for indicators related to tourism

WDIsearch("tourism")

```

Next, we need a list of indicators to work with. The full list of indicators is [here](http://databank.worldbank.org/data/reports.aspx?source=world-development-indicators)

```{r}

#List of indicators to grab
wdilist <- c("NE.TRD.GNFS.ZS",     # Trade (% of GDP)
             "NY.GDP.PCAP.PP.KD",  # GDP per capita, PPP (constant 2005 intl $)
             "NY.GDP.PCAP.KD",     # GDP per capita (constant 2000 US$)
             "NY.GDP.MKTP.KD.ZG",  # GDP growth (annual %)
             "FP.CPI.TOTL.ZG",     # Inflation, consumer prices (annual %)
             "FP.CPI.TOTL",        # Consumer price index (2005 = 100) 
             "NV.AGR.TOTL.ZS",     # Agriculture, value added (% of GDP)
             "NV.IND.MANF.ZS",     # Manufacturing, value added (% of GDP)
             "NV.IND.TOTL.ZS",     # Industry, value added (% of GDP)
             "NV.SRV.TETC.ZS",     # Services, etc., value added (% of GDP)
             "GC.TAX.TOTL.GD.ZS",  # Tax revenue (% of GDP)
             "GC.DOD.TOTL.GD.ZS",  # Central government debt, total (% of GDP)
             "SP.POP.TOTL",        # Population, total
             "SP.URB.TOTL.IN.ZS",  # Urban population (% of total)
             "SP.POP.GROW",        # Population growth (annual %)
             "EN.POP.DNST",        # Population density (people per sq. km of land area)
             "SP.POP.0014.TO.ZS",  # Population ages 0-14 (% of total)
             "MS.MIL.TOTL.P1",     # Armed forces personnel, total
             "MS.MIL.TOTL.TF.ZS",  # Armed forces personnel (% of total labor force)
             "NY.ADJ.DFOR.GN.ZS",  # Adjusted savings: forest depletion (% of GNI)
             "NY.ADJ.DMIN.GN.ZS",  # Adjusted savings: mineral depletion (% of GNI)
             "NY.ADJ.DNGY.GN.ZS",  # Adjusted savings: energy depletion (% of GNI)
             "IT.CEL.SETS.P2",     # Mobile cellular subscriptions (per 100 people)                  
             "IT.NET.USER.P2",     # Internet users (per 100 people)                                
             "SP.DYN.IMRT.IN",
             "SE.SEC.CUAT.LO.ZS")    # Infant mortality rate

#Extract from WDI database: http://data.worldbank.org/data-catalog/world-development-indicators
wd.full<-WDI(indicator=wdilist, country="all", start=1990, end=2015)

#Merge with country info
wd<-left_join(wd.desc,wd.full)

#Make sure region is a character
wd$region<-as.character(wd$region)

#Save file
save(wd,file="wd.RData")

```

Today in class we'll work just with data for Sub-Saharan Africa. The next chunk of code filters for this part of the world and then renames some variables for ease of use. 

```{r}
wd<-wd%>%filter(region=="East Asia & Pacific (all income levels)")

#Rename variables for ease of use
wd$gdp_percap<-wd$NY.GDP.PCAP.PP.KD
wd$infant_mortality<-wd$SP.DYN.IMRT.IN
wd$country2<-as.factor(wd$country)
wd$urban<-wd$SP.URB.TOTL.IN.ZS
wd$youngpop<-wd$SP.POP.0014.TO.ZS
wd$internet<-wd$IT.NET.USER.P2
wd$educ<-wd$SE.SEC.CUAT.LO.ZS
```

Let's start with a simple scatterplot by year. 

```{r}
#Plot GDP per capita by year
g1<-ggplot(data=wd,aes(x=year,y=gdp_percap))
g1<-g1+geom_point()
g1 #Messy!

```

There are several obvious problems. First, a few countries just have much higher GDP per capita than others, which means that for most countries we can't observe change over time. Second, the first question anyone is going to ask is: which country is which? Let's use the `facet_wrap` function to figure out which countries we're talking about.  

```{r}
g2<-g1+facet_wrap(~country,ncol=7)
g2

```

This is our first "small multiples" approach, and its virtues are immediately obvious. A story begins to be told immedietaly about different countries. The discovery of oil in Equatorial Guineau shows up clearly. The missing data for Somalia stands as a stark reminder of the chaos in that country. But, we still have several problems. Most countries look like a flat line, which they're not. They're just swamped by the few high GDP countries. As usual, this calls for a log transform. 

```{r}
g2<-g2+scale_y_continuous(trans="log",breaks=c(0,100,1000,5000,15000,40000))
g2
```

*In Class Exercise*

Create a similar graphic for infant mortaility. 


## Ordering

The next thing I wish I saw a lot more of was ordering by the underlying scale. In our small multiples, Angola comes first. It's rarely of substantive interest to list things alphabetically. Luckily, the `forcats` package has code that can help us. 

```{r}
wd$country<-as.factor(wd$country)
wd<-wd%>%mutate(country=fct_reorder(f=country,x=-infant_mortality))
g3<-ggplot(data=wd,aes(x=year,y=infant_mortality,color=country))
g3<-g3+geom_point()
g3<-g3+scale_y_continuous(trans="log",breaks=c(0,100,1000,5000,15000,40000))
g3<-g3+facet_wrap(~country,ncol=7)
g3
```

We don't need a legend for this graph, so:

```{r}
g3<-g3+theme(legend.position="none")
g3
```

And it seems to me it would look better as a smoothed line:

```{r}
g3<-ggplot(data=wd,aes(x=year,y=gdp_percap,color=country))
g3<-g3+geom_smooth()
g3<-g3+scale_y_continuous(trans="log",breaks=c(0,100,1000,5000,15000,40000))
g3<-g3+facet_wrap(~country,ncol=7)
g3<-g3+theme(legend.position="none")
g3<-g3+xlab("Year")+ylab("GDP per Capita")
## X axis at 45 degree label
g3<-g3+theme(axis.text.x=element_text(angle=45,hjust=1,size=8))

# Change y axis text size
g3<-g3+theme(axis.text.y=element_text(size=10))

## Change facet labels text size
g3<-g3+theme(strip.text.x=element_text(size=8))

g3
```

*Quick exercise: apply the same fixes to your preferred variable*

This graphic now shows us the ranking of countries by GDP, the minimum, maximum and trend for each country, and provides a comparison for the scale of differences. 

## Scatterplots

What explains internet use? Which characteristics of a country might best predict the number of interent users per 100? The graph below shows interent use by country for each year, with a new theme and an interactive element. 

```{r}
g4<-ggplot(data=filter(wd,year>=2001),aes(x=educ,y=internet,color=country))
g4<-g4+geom_point()
g4<-g4+facet_wrap(~year,ncol=5)
g4<-g4+theme_minimal()
g4<-g4+theme(legend.position="none")+xlab("% of Pop<14")+ylab("Internet Users per 100 Pop")
g4

## Interactive Version
g4_plotly<-ggplotly(g4)
g4_plotly
```

*Quick exercise: find another variable to predict interent use. You can use what we already have or go back to the WDI database to find another one*
