---
title: "Mapping in R"
author: "Will Doyle"
output: github_document
---

## Introduction

Mapping is a great way to communicate information. In my experience, people find maps a very easy way to relate to the information that you're trying to provide. Any time you have data that is usefully grouped at a geographic level you should generate some maps to understand spatial patterns. 

We'll need a few new libraries today, including the `maptools` and `maps` libraries. 

```{r}
#############################
# Mapping Data in R
# Will Doyle
# Project start 11/15/2014
#############################

#packages
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(XML)
library(maptools)
library(maps)
library(ggmap)
```

A choropleth map makes use of existing political boundaries. Most maps (state, county) that we're familiar with are actually choropleths. Other types would make use of natrual or analyst-created boundaries. 

For our first choropleth, we're going to work with some data on income inequality. We will use a 50 state map to get the proportion of income that's earned by the top 1 percent of the population. 

```{r}
#State Level Choropleth
#Indebted to:http://trinkerrstuff.wordpress.com/2013/07/05/ggplot2-chloropleth-of-supreme-court-decisions-an-tutorial/

#Inequality data from: http://www.shsu.edu/eco_mwf/inequality.html

ineq<-read.csv("http://www.shsu.edu/eco_mwf/Frank_Top1_v97.csv")

ineq.sub<-ineq[ineq$year==2010,]

ineq.sub$state<-tolower(ineq.sub$state)

states_map<-map_data("state")

g1<-ggplot(ineq.sub, aes(map_id = state,fill=top1)) 
g1<-g1+  geom_map(map = states_map, color ="black") 
g1<-g1+  expand_limits(x = states_map$long, y = states_map$lat) 


g1<-g1+theme(legend.position = "bottom",
        axis.ticks = element_blank(), 
        axis.title = element_blank(), 
        axis.text =  element_blank()
)

g1<-g1+ scale_fill_gradient(low="white", high="orange",name="Percent of Income Earned by Top 1%") 
g1<-g1+  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) 
g1 <- g1 + coord_map("albers",lat0=39, lat1=45)
g1<-g1+theme(legend.position = "bottom")
g1
```

## Top 10 % data

Now, let's do the same thing, but with the 10% data. We'll changes the colorscale and try some different projections. 
```{r}
ineq<-read.csv("http://www.shsu.edu/eco_mwf/Frank_Top10_v97.csv")

ineq.sub<-ineq[ineq$year==2012,]

ineq.sub$state<-tolower(ineq.sub$state)

states_map<-map_data("state")

g1<-ggplot(ineq.sub, aes(map_id = state,fill=top10)) 
g1<-g1+  geom_map(map = states_map, color ="black") 
g1<-g1+  expand_limits(x = states_map$long, y = states_map$lat) 


g1<-g1+theme(legend.position = "bottom",
             axis.ticks = element_blank(), 
             axis.title = element_blank(), 
             axis.text =  element_blank()
)

g1<-g1+  scale_fill_gradient(low="white", high="purple",name="Percent of Income Earned by Top 10% ") 
g1<-g1+  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) 
g1<-g1+theme(panel.border = element_blank(), 
             panel.grid.major = element_blank(),
             panel.grid.minor = element_blank(),
             panel.background=element_blank())
g1

g1 <- g1 + coord_map("albers",lat0=39, lat1=45)
g2 <- g1 + coord_map("mercator")
g3 <- g1 + coord_map("gilbert")
g4 <- g1 + coord_map("stereographic")
g5 <- g1 + coord_map("cylindrical")
g6 <- g1 + coord_map("azequalarea")
```



We can do the same thing with county-level maps. 

```{r}
#County Level Choropleth

#Remember the quick facts data? 


#Check to see if data already in project directory, if so open


  load("quickfacts.Rdata") 
  

  
qf2.sub<-data.frame(qf$fips,qf$coll_grad_pc)
names(qf2.sub)<-c("fips","value")
qf2.sub$fips<-as.numeric(as.character(qf2.sub$fips))
data(county.fips)
qf2.sub<-left_join(qf2.sub,county.fips,by="fips")
qf2.sub$polyname<-as.character(qf2.sub$polyname)


#Get shapefile
county_map<-map_data("county")

county_map$polyname<-paste(as.character(county_map$region),as.character(county_map$subregion),sep=",")

#combine with shapefile
mapdata<-left_join(county_map,qf2.sub,by="polyname")

#Order is important
mapdata <- mapdata[order(mapdata$order), ]

mapdata<-mapdata%>%arrange(order)

#draw map


g2<-ggplot(mapdata, aes(long, lat, group = group,fill=as.numeric(as.character(value)))) +
  geom_polygon(colour = "white", size = 0) + 
  scale_fill_continuous(low = "white",high = "purple",name="Percent W/ College Degree")


g2<-g2+theme(legend.position = "bottom",
             axis.ticks = element_blank(), 
             axis.title = element_blank(), 
             axis.text =  element_blank() )

g2<-g2+  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) 

g2<-g2+theme(panel.border = element_blank(), 
             panel.grid.major = element_blank(),
             panel.grid.minor = element_blank(),
             panel.background=element_blank())

g2


qf2.sub<-data.frame(qf$fips,qf$median_home_val)
names(qf2.sub)<-c("fips","value")
qf2.sub$fips<-as.numeric(as.character(qf2.sub$fips))
data(county.fips)
qf2.sub<-left_join(qf2.sub,county.fips,by="fips")
qf2.sub$polyname<-as.character(qf2.sub$polyname)
qf2.sub$value<-as.numeric(as.character(qf2.sub$value))
qf2.sub$value[qf2.sub$value>250000]<-250000

#Get shapefile
county_map<-map_data("county")

county_map$polyname<-paste(as.character(county_map$region),as.character(county_map$subregion),sep=",")

#combine with shapefile
mapdata<-left_join(county_map,qf2.sub,by="polyname")

#Order is important
mapdata <- mapdata[order(mapdata$order),]

#draw map

g2<-ggplot(mapdata, aes(long, lat, group = group,fill=as.numeric(as.character(value)))) +
  geom_polygon(colour = "black", size = 0.2) + 
  scale_fill_continuous(low = "white",high = "purple",name="Median Home Value")


g2<-g2+theme(legend.position = "bottom",
             axis.ticks = element_blank(), 
             axis.title = element_blank(), 
             axis.text =  element_blank() )

g2<-g2+  guides(fill = guide_colorbar(barwidth = 10, barheight = .5)) 


g2




```

The `blscraper` package automates this process, while the `tigris` package provides up-to-date shape files. The `leaflet` package can make interactive graphics. This is kind of crazy, to be honest. 

```{r}

## From: https://cran.r-project.org/web/packages/blscrapeR/vignettes/Mapping_BLS_Data.html

# Leaflet map
library(blscrapeR)
library(tigris)
library(leaflet)
map.shape <- counties(cb = TRUE, year = 2015)
df <- get_bls_county()

df<-df%>%mutate(employ_rate=employed/labor_force)

# Slice the df down to only the variables we need and rename "fips" colunm
# so I can get a cleaner merge later.
df <- subset(df, select = c("employ_rate", "fips"))
colnames(df) <- c("employ_rate", "GEOID")



# Merge df with spatial object.
leafmap <- geo_join(map.shape, df, by="GEOID")

# Format popup data for leaflet map.
popup_dat <- paste0("<strong>County: </strong>", 
                    leafmap$NAME, 
                    "<br><strong>Value: </strong>", 
                    leafmap$employ_rate)

pal <- colorQuantile("PuOr", NULL, n = 10)

# Render final map in leaflet.
lmap<-leaflet(data = leafmap) %>% addTiles() %>%
    addPolygons(fillColor = ~pal(employ_rate), 
                fillOpacity = 0.8, 
                color = "#BDBDC3", 
                weight = 1,
                popup = popup_dat)



```

